---
title: "alpscarf"
author: "Chia-Kai Yang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{alpscarf tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
## Initial Setup
Import the library we suggest to use and the package _alpscarf_.
```{r setup, echo = FALSE}
library(tidyverse)
import::from(cowplot, plot_grid)
import::from(magrittr, "%<>%")
library(alpscarf)
```

## Data Import
Import example data. The synthetic dataset _example_eye_movement_data.RData_ includes __AOI visit sequences__ for 3 participants (P1 - P3). The dataset _aoi_sequence.RData_ contains the __expected visit order__ of each AOI, as well as the __color association__ of AOIs.
```{r get dataset}
# specify dataset name & path
dataset_path = "../data"
example_dataset <- file.path(dataset_path, "example_eye_movement_data.RData")
expected_aoi_sequence <- file.path(dataset_path, "aoi_sequence.RData")

# load data  
load(example_dataset)
load(expected_aoi_sequence)

# put AOIs in sorted order
aoi_names_pages_seq %<>%
  arrange(AOI_order)
```

The __AOI visit sequences__ contains at least 3 columns: "p_name" "AOI" "dwell_duration"  
```{r}
head(eye_movement_data_systhetic)
```

The __expected visit order__ contains at least two columns: "AOI" and "AOI_order".
```{r}
head(aoi_names_pages_seq)
```
In above example dataset, we merge the __expected visit order__ with and __color association__, resulting the an extra column ```color```.

## Generate Alpscarf
When the data is ready, users can generate Alpscarf in following steps:

###1. Define colors for AOIs

The association between AOIs and colors MUST be an 1-to-1 mapping; i.e., n AOIs require n colors. An exmaple of such mapping is included in the dataset _aoi_sequence.RData_.
```{r color association}
# define colors
my_palette <- 
  aoi_names_pages_seq$color[]
```

###2. Translate visit sequence into bar height/width 

Two funcitons are involved in this step:

* _alpscarf_height_trans_: the function calculates the confomity/revisiting scores and translate them into bar height
* _alpscarf_width_trans_: the function calculates the position of bars where bar width equals dwell duration (in logarithmic scale)
```{r spec translation}
# generate Alpscarf dataset
eye_movement_data_systhetic_alp_df <- 
  eye_movement_data_systhetic %>%
  alpscarf_add_height(., aoi_names_pages_seq, height_mode = "linear") %>%
  alpscarf_add_width()
```
Note that some parameters are specifiable in the function _alpscarf_height_trans_. Users can choose from linear mode (```height_mode = "linear"```) and exponential mode (```height_mode = "exponential"```), which controls the contrast of mountain height (i.e., make high mountains even higher, and vice versa). In exponential mode, users can also specify the base (```base_factor```) and the scale (```scale_factor```) of exponent.

###3. Generate Alpscarf plot

Now the dataset is ready for generating Alpscarf. In this example, we generate one Alpscarf per participant.

###Example 1: Alpscarf in normalized view
```{r alpscarf plot}
# initialise the list storing Alpscarfs, one plot per participant
lsa_scarf_vis <- list()
# specify plot height
plot_height <- max(eye_movement_data_systhetic_alp_df$seq_bar_length)

for(a_p_name in unique(eye_movement_data_systhetic_alp_df$p_name)){
  df_p <-
    eye_movement_data_systhetic_alp_df %>%
    filter(p_name == a_p_name) 
  
  # for indexing
  a_p_nr <- 
    strsplit(a_p_name,"P")[[1]][2] %>%
    as.numeric()
  
  # specify the color association of AOIs
  aoi_name_in_order <- 
    aoi_names_pages_seq$AOI[]
  df_p$AOI <- factor(df_p$AOI, levels = aoi_name_in_order)
  
  # Alpscarf plot generation
  lsa_scarf_vis[[a_p_nr]] <- alpscarf_plot(df_p, my_palette, focus_mode = "transition", plot_type = "alpscarf", ymax = plot_height, NORMALIZED_VIEW = TRUE, title = a_p_name)
}

# plot Alpscarf for all participants
plot_grid(lsa_scarf_vis[[1]], lsa_scarf_vis[[2]], lsa_scarf_vis[[3]], ncol = 1)
```

Notice that, ```ymax``` specify the height of each plot. To make all Alpscarf comparable, generating Alpscarfs in the same height is highly recommended.

In Example 1, the generate plots are: 

* __Alpscarf__ (```plot_type = "alpscarf"```): users can also generate traditional scarf plots (```plot_type = "scarf"```).
* __transition-focus__ mode (```focus_mode = "transition"```): Users can also generate Alpscarf in __duration-focus__ mode (```focus_mode = "duration"```)
* __normalized view__ (```NORMALIZED_VIEW = TRUE```): all generated plots are of the same length. Users can also generate unnormalized view (```NORMALIZED_VIEW = FALSE```) as shown in the next example.

###Example 2: Alpscarf in unnormalized view
```{r alpscarf plot in unnormalized view}
# initialise the list storing Alpscarfs, one plot per participant
lsa_scarf_vis <- list()
# specify plot height
plot_height <- max(eye_movement_data_systhetic_alp_df$seq_bar_length)
# specify plot width range; used in unnormalized view
max_nr_transitions <- 
  eye_movement_data_systhetic_alp_df$trial %>%
  max()
max_sum_dwell_duration <-
  eye_movement_data_systhetic_alp_df %>% 
  group_by(p_name) %>% 
  summarise(total = sum(dwell_duration)) %>% 
  select(total) %>% 
  max()

for(a_p_name in unique(eye_movement_data_systhetic_alp_df$p_name)){
  df_p <-
    eye_movement_data_systhetic_alp_df %>%
    filter(p_name == a_p_name) 
  
  # for indexing
  a_p_nr <- 
    strsplit(a_p_name,"P")[[1]][2] %>%
    as.numeric()
  
  # specify the color association of AOIs
  aoi_name_in_order <- 
    aoi_names_pages_seq$AOI[]
  df_p$AOI <- factor(df_p$AOI, levels = aoi_name_in_order)
  
  # Alpscarf plot generation
  # Alpscarf plot generation
    lsa_scarf_vis[[a_p_nr]] <- alpscarf_plot(df_p, my_palette, focus_mode = "transition", plot_type = "alpscarf", ymax = plot_height, NORMALIZED_VIEW = FALSE, max_nr_transitions = max_nr_transitions, max_sum_dwell_duration = max_sum_dwell_duration, title = a_p_name)
}

# plot Alpscarf for all participants
plot_grid(lsa_scarf_vis[[1]], lsa_scarf_vis[[2]], lsa_scarf_vis[[3]], ncol = 1)
```

In addition to setting ```NORMALIZED_VIEW = FALSE```, unnormalized view requires two extra arguments: ```max_nr_transitions``` and ```max_sum_dwell_duration```, which define the maximun length of plots in __transition-focus__ mode and __duraiton-focus__ mode respectivelty.


###4. (bonus) Distance measures of AOI visit sequences

__This part is NOT necessary to generate Alpscarf plot__

Additionally, users can get a few __descriptive stats__ (e.g., order_following, rereading ratio), and __edit distance__ between each AOI sequence and the expected visit order. 
```{r edit distance}
# generate stats of Alpscarf; NOT required in generating Alpscarf.
stat_table_df <- alpscarf_stats(eye_movement_data_systhetic_alp_df, aoi_names_pages_seq)

head(stat_table_df)
```
